Probabilistic Catalogs for Crowded Stellar Fields

### 2013-01-31 paper/ms.tex
Various edits, including some softer language

\begin{abstract}
We present a probabilistic (Bayesian) method for producing catalogs
from images of stellar fields. The method is capable of inferring the
number of sources $N$ in the image and can also handle the
challenges introduced by overlapping sources. The luminosity
function of the stars can also be inferred even when the precise luminosity
of each star is uncertain. The method is implemented using advanced
Markov Chain Monte Carlo (MCMC) techniques including Reversible Jump
and Nested Sampling.
This is in contrast with standard techniques which typically produce a single catalog,
potentially underestimating the uncertainties in any study of the stellar
population and discarding information about sources at or below the detection limit.
While standard approaches are often successful, they are more likely to fail
when the image contains many overlapping sources.

The computational feasibility of the method is
demonstrated on simulated data where the luminosity function of the stars is a
broken power-law. The parameters of the luminosity function can be recovered
with moderate uncertainties. We compare the results obtained from our method
with those obtained from the \sex~software and find that the latter significantly
underestimates the number of stars in the image and leads to incorrect
inferences about the luminosity function of the stars.
\end{abstract}

@@

Telescopes don't make catalogs \citep{2011EAS....45..351H}, they
measure the intensity field.  Viewed through the lens of probabilistic
inference, the goals of astronomy are to take the information in the
telescope-generated records of the intensity field and deliver it to
the quantities of astronomical interest with as little loss as
possible.  Insertion of a catalog-generation step in the inference
pipeline between the raw imaging and the final astrophysical analyses
is typically very lossy; the hard decisions of catalog making destroy
information.  Probability theory suggests that it will be much less
lossy to pass forward not a catalog but a probabilisitic
description of all the catalogs that could be consistent with the
imaging---a posterior probability in (enormously large) catalog space.
This article represents an attempt at this ambitous goal in the specific
situation where the only objects in the field are stars.

Beyond these extremely general concerns, there are practical issues:
Standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, uncertainty should be
taken into account. Instead of simply {\it estimating} the position and flux of
each object in the image, we should fully describe the fact that sometimes we
are not certain of the position and flux of each source. This uncertainty
should then be propagated into any inferences about the stellar population.

@@

\section{Bayesian Inference}
To quantitatively model uncertainties and transform noise in observed data
into uncertainties in parameters of interest, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha, mackay}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta).
\end{equation}
Note that this is a very concise notation \citep{hogg} and should be read
as ``the probability distribution for $\theta$''.
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}:
\begin{equation}
p(x|\theta).
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the space of possible data sets, one probability
distribution for each possible value
of $\theta$. Note that the choice of
the sampling distribution is also an assumption about prior knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way. Without this prior knowledge, learning is
impossible: there has to be some relationship between the parameters and the
data, otherwise new data tells you about nothing but itself.

@@

To proceed with the model for inferring catalogs (stellar positions and fluxes)
from image data, we must
specify a definite hypothesis space and choices for the prior distribution
and the sampling distribution. These choices are presented and discussed in
Section~\ref{sec:model}. In Section~\ref{sec:mcmc} we briefly discuss our
MCMC implementation. Section~\ref{sec:simulated_data} describes the tests we
carried out on simulated data, and a comparison with \sex~results is presented
in Section~\ref{sec:sex}. We conclude in Section~\ref{sec:conclusion}.
%Throughout this paper we make a number of simplifications
%for the purposes of testing the computational feasibility and performance of
%the method. In real applications, some of these assumptions (such as known PSF
%and noise variance) may be relaxed.

@@

\subsection{The Prior}
The prior probability distribution for the unknown parameters can be factorized
using the product rule of probability theory.
With a variety of independence assumptions, the prior
can be factorized as:
\begin{eqnarray}
p(\theta) = p(\beta)p(N|\beta)\prod_{i=1}^N p(x_i, y_i)
p(f_i | \beta)
\end{eqnarray}
Here, we have assumed that the luminosity function does not depend on position.
Finally, the fluxes of the stars come independently from
a common distribution. If we knew the luminosity
function of the stars, then the location and flux of a particular star would
not tell us anything about the location and flux of another star. Really, this
is just a way of implementing exchangeability of the stars.

@@

\subsection{The Sampling Distribution}
The sampling distribution is a probabilistic model for the process that
generates the data; it describes the probability distribution we would use
to predict the data if we happened to know the true catalog. In our case,
the data will be an $m \times n$ array of pixel intensities $I$:
\begin{eqnarray}
\{I_{ij}\}
\end{eqnarray}
where the central position of each pixel is:
\begin{eqnarray}
\{X_{ij}, Y_{ij}\}.
\end{eqnarray}
The image is assumed to be a noisy version of the true underlying intensity
field. Thus, we need
a prescription for simulating an image $\{I_{ij}\}$ from a catalog $\theta$:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N,
\left\{f_i\right\}_{i=1}^N\right\}.
\end{eqnarray}
If we knew the true catalog, we could compute the
``mock image'' we would expect to see
in the absence of noise. This mock image (defined at every point on the sky)
is given by:
\begin{eqnarray}
\mathcal{M}(x, y) &=& \sum_{i=1}^N f_i \mathcal{P}(x - x_i, y - y_i)
\end{eqnarray}
where $\mathcal{P}$ is the pixel-convolved point spread function (PSF).
Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians with widths $s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of each pixel is determined by the brightness of the sky and the
brightness of the mock image in that pixel:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

@@

\subsection{Test Case 2}
Test Case 2 is more challenging than Test Case 1 because the image contains
more stars. This increases the size of the computational task in a number of
ways: firstly, there will be more unknown parameters to infer, so the
convergence of the MCMC algorithm will take longer. Secondly, the time taken
to compute the predicted image from a catalog (in order to evaluate the
likelihood) is longer because of the larger number of stars. Using DNS, some
samples from the posterior distribution can be obtained in about a day on a
modern PC. However, this problem does not have a first-order phase transition
like Test Case 1. Therefore, it is tractable to run standard Metropolis-Hastings
MCMC targeting the posterior distribution. In fact, this can be more efficient
because the degree of compression from the prior distribution to the posterior
distribution is very large ($\sim e^{3000}$) in this case.

>>>>

\begin{abstract}
We present a probabilistic (Bayesian) method for producing catalogs
from images of stellar fields. The method is capable of inferring the
number of sources $N$ in the image and can also handle the
challenges introduced by overlapping sources. The luminosity
function of the stars can also be inferred even when the precise luminosity
of each star is uncertain. The computational feasibility of the method is
demonstrated on simulated data where the luminosity function of the stars is a
broken power-law. We compare the results obtained from our method
with those obtained from the \sex~software and find that the latter ...
\end{abstract}

@@

Telescopes don't make catalogs \citep{2011EAS....45..351H}, they
measure the intensity field.  Viewed through the lens of probabilistic
inference, the goals of astronomy are to take the information in the
telescope-generated records of the intensity field and deliver it to
the quantities of astronomical interest with as little loss as
possible.  Insertion of a catalog-generation step in the inference
pipeline between the raw imaging and the final astrophysical analyses
is potentially very lossy; the hard decisions of catalog making destroy
information.  Probability theory suggests that it may be less
lossy to pass forward not a catalog but a probabilisitic
description of all the catalogs that could be consistent with the
imaging---a posterior probability in (enormously large) catalog space.
This article represents an attempt at this ambitous goal in the specific
situation where the only objects in the field are stars or other point sources.

Beyond these extremely general concerns, there are practical issues:
Standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, the uncertainty about
the existence and properties of the stars can be significant and should be
propagated into any inferences about the stellar population.

@@

The remainder of this paper is structured as follows. In Section~\ref{sec:bayes}
we give a brief overview of Bayesian inference. In Section~\ref{sec:model} we
discuss the model assumptions we make in our method. In Section~\ref{sec:mcmc}
we briefly discuss our
MCMC implementation. Section~\ref{sec:simulated_data} describes the tests we
carried out on simulated data, and a comparison with \sex~results is presented
in Section~\ref{sec:sex}. We conclude in Section~\ref{sec:conclusion}.

\section{Bayesian Inference}\label{sec:bayes}
To quantitatively model uncertainties and transform noise in observed data
into uncertainties in parameters of interest, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha, mackay}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta).
\end{equation}
Note that this is a very concise notation \citep{hogg} and should be read
as ``the probability distribution for $\theta$''.
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}:
\begin{equation}
p(x|\theta).
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the space of possible data sets, one probability
distribution for each possible value
of $\theta$. Note that the choice of
the sampling distribution is also an assumption about prior knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way \citep{caticha}. Without this prior
knowledge, learning is impossible: there has to be some relationship between
the parameters and the data, otherwise new data tells you about nothing but
itself.

@@

To proceed with the model for inferring catalogs
from image data, we must
specify a definite hypothesis space and choices for the prior distribution
and the sampling distribution. These choices are presented and discussed in
Section~\ref{sec:model}.
%Throughout this paper we make a number of simplifications
%for the purposes of testing the computational feasibility and performance of
%the method. In real applications, some of these assumptions (such as known PSF
%and noise variance) may be relaxed.

@@

\subsection{The Prior}
The prior probability distribution for the unknown parameters can be factorized
using the product rule of probability theory.
With a variety of independence assumptions, the prior
can be factorized as:
\begin{eqnarray}
p(\theta) = p(\beta)p(N|\beta)\prod_{i=1}^N p(x_i, y_i)
p(f_i | \beta)
\end{eqnarray}
Here, we have assumed that the luminosity function does not depend on position.
Finally, the fluxes of the stars come independently from
a common distribution. If we knew the luminosity
function of the stars, then the location and flux of a particular star would
not tell us anything about the location and flux of another star. Really, this
is just a way of implementing exchangeability of the stars, and is often
called a {\it hierarchical model}.

@@

\subsection{The Sampling Distribution}
The sampling distribution is a probabilistic model for the process that
generates the data; it describes the probability distribution we would use
to predict the data if we happened to know the true catalog. In our case,
the data will be an $m \times n$ array of pixel intensities $I$:
\begin{eqnarray}
\{I_{ij}\}
\end{eqnarray}
where the central position of each pixel is:
\begin{eqnarray}
\{X_{ij}, Y_{ij}\}.
\end{eqnarray}
The image is assumed to be a noisy version of the true underlying intensity
field. Thus, we need
a prescription for simulating an image $\{I_{ij}\}$ from a catalog $\theta$:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N,
\left\{f_i\right\}_{i=1}^N\right\}.
\end{eqnarray}
If we knew the true catalog, we could compute the
``mock image'' we would expect to see
in the absence of noise. This mock image (defined at every point on the sky)
is given by:
\begin{eqnarray}
\mathcal{M}(x, y) &=& \sum_{i=1}^N f_i \mathcal{P}(x - x_i, y - y_i)
\end{eqnarray}
where $\mathcal{P}$ is the pixel-convolved point spread function (PSF).
Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians with widths $s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of each pixel is determined by the brightness of the sky and the
brightness of the mock image in that pixel:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image. This dependence of the
noise variances on the model intensity arises as a result of the Poissonian
nature of photon counts, but allows for the fact that a ``sky'' background may
have already been subtracted from the image.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

@@

\subsection{Test Case 2}
Test Case 2 is more challenging than Test Case 1 because the image contains
more stars. This increases the size of the computational task in a number of
ways: firstly, there will be more unknown parameters to infer, so the
convergence of the MCMC algorithm will take longer. Secondly, the time taken
to compute the predicted image from a catalog (in order to evaluate the
likelihood) is longer because of the larger number of stars. Using DNS, some
samples from the posterior distribution can be obtained in about a day on a
modern PC. However, this problem does not have a first-order phase transition
like Test Case 1. Therefore, it is tractable to run standard Metropolis-Hastings
MCMC targeting the posterior distribution. In fact, this can be more efficient
because the efficiency of Nested Sampling methods is sensitive to the ``information'',
or the amount of prior mass occupied by the posterior distribution. In this case,
because of the large number of parameters, the information is very large, about
3000 nats.
